---
title: 'ENV872 Final Project'
author: "Keanu Valibia" 
date: "Spring 2024"
output:
  pdf_document: default
  html_document: default
---
\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
# Set your working directory

setwd("~/R/R Projects/Spring24-ENV872-Final")

# Load your packages

library(tidyverse)
library(lubridate)
library(dplyr)
library(ggplot2)
library(here)
#install.packages("wbstats")
library(wbstats)

# Set your ggplot theme

myTheme <- theme_classic(base_size = 11) +
  theme(axis.text = element_text(color = "black"),
        axis.line = element_line(color = "black"),
        panel.background = element_rect(fill ="#EDE6E3"),  
        panel.grid.major = element_line(color = "#36382E",  
                                        linetype = "dotted"),  
        plot.title = element_text(size = 15),  
        axis.title.x = element_text(size = 13),  
        axis.title.y = element_text(size = 13),
        legend.position = "right")

theme_set(myTheme)

# Load your datasets

#EG.USE.PCAP.KG.OE: Energy use per country in terms of kg of oil equivalent per capita
#SH.DYN.MORT: Child mortality rates under the age of 5 per 1,000 live births
#EN.ATM.GHGT.KT.CE: Total greenhouse gas emissions as kilotons of CO2 equivalent
#EG.FEC.RNEW.ZS: Estimated usage of renewable energies as a percentage of total final energy consumption

#Import data from World Bank API. Rename fields to make them more intuitive
energyUse_mort <- wb_data(country = c("United Arab Emirates","United Kingdom","Argentina", "Japan", "Italy", "United States"),
                          indicator=c("EG.USE.PCAP.KG.OE","SH.DYN.MORT","EN.ATM.GHGT.KT.CE","EG.FEC.RNEW.ZS"), start_date = 1990, end_date = 2020) %>%
  rename(Energy_Use = EG.USE.PCAP.KG.OE, Mortality_Rate = SH.DYN.MORT, Total_GHG_Emissions = EN.ATM.GHGT.KT.CE, Renewable_Consump = EG.FEC.RNEW.ZS)

#Export data as a .csv file.
write.csv(energyUse_mort, "World_Bank_EnergyUse_Mortality.csv")

```


# Rationale and Research Questions

This data project utilizes data from the World Bank spanning child mortality rates, total energy consumption, renewable energy consumption, and total greenhouse gas (ghg) emissions. The purpose of this lab is to understand any potential correlation between mortality rates and ghg emissions / energy consumption. Thus, the questions for this project would be:

Question 1: H0 = There is no observable effect of energy consumption, renewable energy consumption, and total ghg emissions on mortality rates (H0 = 0).

Question 2: Ha = There is observable effect of energy consumption, renewable energy consumption, and total ghg emissions on mortality rates (H0 != 0).

The World Bank API is used to pull data sources across the four mentioned variables. This data is used as a reliable source of data that spans across decades of data collection. Mortality rates, (specifically those for under the age of 5 per 1,000 deaths) are used as the independent variable as a proxy for development of a nation. GHG emissions are used as a very rough proxy for not just how pollutant a country may be, but also for level of development. Renewable energy consumption is used as a proxy for how sustainable a country may be, while total energy consumption may be used as another proxy for how productive or developed a country is.

\newpage

# Dataset Information

Table: WorldBank Dataset Information

| Item     | Value                               |
| ---:     | :---                                |
| Source   | World Bank API                      |
| Date     | 1990 - 2020                         |
| Filename | World_Bank_EnergyUse_Mortality.csv  |

Table: WorldBank Dataset Information - Fields

| Item                | Value                                                                                    |
| ---:                | :---                                                                                     |
| iso2c               | 2 character country acronym                                                              |
| iso2c               | 2 character country acronym                                                              |
| country             | Country name                                                                             |
| date                | Year row data was collected                                                              |
| Renewable_Consump   | Estimated usage of renewable energies as a percentage of total final energy consumption  |
| Energy_Use          | Energy use per country in terms of kg of oil equivalent per capita                       |
| Total_GHG_Emissions | Total greenhouse gas emissions as kilotons of CO2 equivalent                             |
| Mortality_Rate      | Child mortality rates under the age of 5 per 1,000 live births                           |

\newpage

# Exploratory Analysis 

```{r}

#Review first few values of dataframe
glimpse(energyUse_mort)
```

```{r Data Summary, warning=FALSE, message=FALSE}

#Fetch dimensions of df
dim(energyUse_mort)

#Fetch summary statistics of df across fields
summary(energyUse_mort)

```

```{r Mortality Distribution Analysis, fig.cap = "Mortality Distribution Analysis", fig.align='left', fig.height=3.6}

#Create distrubtion plots of independent and dependent variables and review for heteroskedasticity.

mortality_dist <- ggplot(energyUse_mort, aes(x = Mortality_Rate)) +
  geom_histogram() +
  labs(title = "Energy Consumption Distribution",
       caption = "Per Country",
       color = "Country") +
  xlab("Mortality Rate") +
  ylab("Frequency")

mortality_dist
```

```{r GHG Distribution Analysis, fig.cap = "GHG Distribution Analysis", fig.align='left', fig.height=3.6}
ghg_dist <- ggplot(energyUse_mort, aes(x = Total_GHG_Emissions)) +
  geom_histogram() +
  labs(title = "GHG Distribution",
       caption = "Per Country",
       color = "Country") +
  xlab("Mortality Rate") +
  ylab("Frequency")

ghg_dist
```

```{r Total Energy Use Distribution Analysis, fig.cap = "Total Energy Use Distribution Analysis", fig.align='left', fig.height=3.6}
EnergyUse_dist <- ggplot(energyUse_mort, aes(x = Energy_Use)) +
  geom_histogram() +
  labs(title = "Total Energy Use",
       caption = "Per Country",
       color = "Country") +
  xlab("Mortality Rate") +
  ylab("Frequency")

EnergyUse_dist
```

```{r Renewable Distribution Analysis, fig.cap = "Renewable Distribution Analysis", fig.align='left', fig.height=3.6}
Renewable_Consump_dist <- ggplot(energyUse_mort, aes(x = Renewable_Consump)) +
  geom_histogram() +
  labs(title = "Renewable Energy Use Distribution",
       caption = "Per Country",
       color = "Country") +
  xlab("Mortality Rate") +
  ylab("Frequency")

Renewable_Consump_dist
```

```{r Mortality Logged Distribution Analysis, fig.cap = "Mortality Logged Distribution Analysis", fig.align='left', fig.height=3.6}

#Create logged versions of histograms and review for normalized graphs
mortality_dist <- ggplot(energyUse_mort, aes(x = log(Mortality_Rate))) +
  geom_histogram() +
  labs(title = "Logged Mortality Rate Distribution",
       caption = "Per Country",
       color = "Country") +
  xlab("Mortality Rate") +
  ylab("Frequency")

mortality_dist
```

```{r GHG Logged Distribution Analysis, fig.cap = "GHG Logged Distribution Analysis", fig.align='left', fig.height=3.6}
ghg_dist <- ggplot(energyUse_mort, aes(x = log(Total_GHG_Emissions))) +
  geom_histogram()  +
  labs(title = "Logged GHG Distribution",
       caption = "Per Country",
       color = "Country") +
  xlab("Mortality Rate") +
  ylab("Frequency")

ghg_dist
```

```{r Total Energy Logged Distribution Analysis, fig.cap = "Total Energy Distribution Analysis", fig.align='left', fig.height=3.6}
EnergyUse_dist <- ggplot(energyUse_mort, aes(x = log(Energy_Use))) +
  geom_histogram() +
  labs(title = "Total Energy Used Logged Distribution",
       caption = "Per Country",
       color = "Country") +
  xlab("Mortality Rate") +
  ylab("Frequency")

EnergyUse_dist
```

```{r Renewable Energy Logged Distribution Analysis, fig.cap = "Renewable Energy Distribution Analysis", fig.align='left', fig.height=3.6}
Renewable_Consump_dist <- ggplot(energyUse_mort, aes(x = log(Renewable_Consump))) +
  geom_histogram() +
  labs(title = "Renewable Energy Logged Distribution",
       caption = "Per Country",
       color = "Country") +
  xlab("Mortality Rate") +
  ylab("Frequency")

Renewable_Consump_dist
```

```{r Renewable Energy Scatterplot, fig.cap = "Renewable Energy Scatterplot", fig.align='left', fig.height=3.6}

#Create scatterplots to review changes in independent variables over time.

renewable_consump_time <- ggplot(energyUse_mort, aes(x = date, y = Renewable_Consump, color = country)) +
  geom_point() + 
  labs(title = "Renewable Energy Consumption Over Time",
       caption = "Per Country",
       color = "Country") +
  xlab("Year") +
  ylab("Renewable Energy Used (% of Total Use")

renewable_consump_time
```

```{r Total Energy Use Scatterplot, fig.cap = "Total Energy Use Scatterplot", fig.align='left', fig.height=3.6}
energy_consump_time <- ggplot(energyUse_mort, aes(x = date, y = Energy_Use, color = country)) +
  geom_point() + 
  labs(title = "Energy Consumption as kg of Oil Equivalent per Capita",
       caption = "Per Country",
       color = "Country") +
  xlab("Year") +
  ylab("Energy Consumption")

energy_consump_time
```

```{r GHG Scatterplot, fig.cap = "GHG Scatterplot", fig.align='left', fig.height=3.6}
ghg_emissions_time <- ggplot(energyUse_mort, aes(x = date, y = Total_GHG_Emissions, color = country)) +
  geom_point() + 
  labs(title = "Total GhG Emissions Over Time (kt of CO2 Equivalent)",
       caption = "Per Country",
       color = "Country") +
  xlab("Year") +
  ylab("GhG Emissions")

ghg_emissions_time
```

```{r Mortality Scatterplot, fig.cap = "Mortality Scatterplot", fig.align='left', fig.height=3.6}
mortality_time <- ggplot(energyUse_mort, aes(x = date, y = Mortality_Rate, color = country)) +
  geom_point() + 
  labs(title = "Mortality Rate (Under 5 per 1,000 Births) per Country Over Time",
       caption = "Per Country",
       color = "Country") +
  xlab("Year") +
  ylab("Mortality Rate")

mortality_time 
```

\newpage

# Analysis

```{r multiple regression setup}

#Use AIC function to review which independent variables can be used for multiple regression.
AIC <- lm(data = energyUse_mort, log(Mortality_Rate) ~ Renewable_Consump + Energy_Use + Total_GHG_Emissions)
AIC

step(AIC)

```

```{r multiple regression}

# Run multiple regressions
AICmodel <- lm(data = energyUse_mort,  log(Mortality_Rate) ~ Renewable_Consump + Energy_Use + Total_GHG_Emissions)
summary(AICmodel)

AICmodel
```

## Question 1: H0 = There is no observable effect of energy consumption, renewable energy consumption, and total ghg emissions on mortality rates (H0 = 0).

## Question 2: Ha = There is observable effect of energy consumption, renewable energy consumption, and total ghg emissions on mortality rates (H0 != 0).


\newpage

# Summary and Conclusions
Although there is enough evidence to reject the null hypothesis, the regression coefficients seem to oddly indicate that increases in both renewable energy consumption and total energy use seems to correlate to increased mortality rates, while a decrease in total greenhouse gas emissions seem to correlate with a decrease in mortality rates for every increase in total greenhouse gas emissions.

All of these regression coefficients are statistically significant, with renewable consumption and total ghg emissions significant at the 5 percent level, and energy use significant at the 1 percent level.

However, there are a few things to point out. Quantitatively, the R-Squared is only 0.056, which means that only about 5.6% of variance in the model is explaned by this model, indicating drastic underfit of data. Additional data will be needed to raise this R-Squared to an acceptable level without overfitting the data.

Qualitatively, there are quite a few exogenous factors missing from the model that may help explain or even drastically change these results. Perhaps increases in energy consumption and energy use may correlate to higher mortality rates simply because of larger populations. Perhaps there are other factors within individual countries' economies that may explain higher mortality rates such as crime rates, poverty rates, etc. Negative correlation between mortality rates and ghg emissions may potentially be explained better by combination of factors including total population, total gdp, etc. Countries with higher greenhouse gas emissions may emit so much more than less developed countries that the data may be skewed. Industrial countries that pollute much more than smaller, less-developed countries tend to have more advanced infrastructure to support advanced medical facilities, etc. Thus, mortality rates cannot be explained soley by these three factors alone.

github repository link: https://github.com/ItsTheKGV/Spring24-ENV872-Final/tree/main

\newpage

# References
World Bank Data
Mortality Rate: https://data.worldbank.org/indicator/SH.DYN.MORT
Renewable Energy Consumption: https://data.worldbank.org/indicator/EG.FEC.RNEW.ZS?view=chart
Energy Use: https://data.worldbank.org/indicator/EG.USE.PCAP.KG.OE?view=chart
Total Greenhouse Gas Emissions: https://data.worldbank.org/indicator/EN.ATM.GHGT.KT.CE?view=chart